Evolution of pre-copulatory and post-copulatory strategies of inbreeding avoidance

### 2017-05-03 pre_post_biorxiv.tex
Final formatting of biorXiv TEX file

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

We model evolution of polyandry, and of pre-copulatory and post-copulatory inbreeding avoidance strategies (hereafter simply 'inbreeding strategies' because the model did not preclude evolution of inbreeding preference), by tracking the dynamics of alleles underlying reproductive strategies enacted by interacting females and males within a small focal population. We thereby track evolutionary dynamics given internally consistent patterns of relatedness caused by non-random mating and allowing for effects of mutation, gene flow, drift, and selection. Each individual has 10 diploid loci, and therefore 20 alleles, underlying each of three reproductive strategy traits (i.e., 30 diploid loci in total): tendency for polyandry ($P_{a}$, i.e., `polyandry' alleles), pre-copulatory inbreeding strategy ($M_{a}$, i.e., `mating' alleles), and post-copulatory inbreeding strategy ($F_{a}$, i.e., `fertilisation' alleles). All alleles can take the value of any real number \cite[i.e., a continuum-of-alleles model;][]{Kimura1965, Lande1976, Reeve2000, Bocedi2014} and combine additively to affect genotypes ($P_{g}$, $M_{g}$, and $F_{g}$) and phenotypes ($P_{p}$, $M_{p}$, and $F_{p}$) for tendency for polyandry, pre-copulatory inbreeding strategy, and post-copulatory inbreeding strategy, respectively, whereby an individual's genotypic values $P_{g}$, $M_{g}$, and $F_{g}$ equal the sum of its 20 alleles for each trait. In overview, each generation proceeds with females paying costs and expressing polyandry, mating, and fertilisation. Offspring inherit a randomly sampled allele from each parent at each locus with no physical linkage. Alleles can then mutate and offspring express inbreeding depression in viability. Immigrants arrive in the population and density regulation limits population growth. We record the population pedigree and directly calculate the coefficient of kinship ($k$) between all potential mates in each generation (defined as the probability that two homologous alleles will be identical-by-descent), allowing individual pre-copulatory and post-copulatory inbreeding strategies to be enacted. Key individual traits, parameter values, and variables are described in Table 1.

@@

Phenotypic values of $P_{p}$, $M_{p}$, and $F_{p}$ each have costs that are assumed to independently increase the probability that a focal female will die before mating (see Discussion for consideration of other costs). The probabilities of pre-mating mortality due to the costs of polyandry ($c_{P}$), pre-copulatory inbreeding strategy ($c_{M}$), and post-copulatory inbreeding strategy ($c_{F}$) are $P_{p} \times c_{P}$, $|M_{p}| \times c_{M}$, and $|F_{p}| \times c_{F}$, respectively. Here $|M_{p}|$ and $|F_{p}|$ are the absolute values of $M_{p}$ and $F_{p}$, respectively. Absolute values are used for applying costs to inbreeding avoidance strategies because both negative and positive $M_{p}$ and $F_{p}$ values could potentially arise and affect the degree of inbreeding, representing inbreeding avoidance and inbreeding preference, respectively (see details of mating and fertilisation strategies below). In contrast, only positive values of $P_{p}$ cause females to be polyandrous (see details of polyandry below). Overall, because generations are non-overlapping, a female's probability of total reproductive failure increases linearly with the phenotypic value of each trait.

@@

\subsection*{Polyandry}

After costs are realised, each remaining female chooses $N_{males}$ males to mate with, where $N_{males}$ is calculated by sampling from a Poisson distribution such that $N_{males} = Poisson(P_{p}) + 1$. This ensures that all surviving females choose at least one mate, and causes some stochastic variation around the expected mean $N_{males}$ of $P_{p}+1$. A female's value of $P_{p}$ depends on the sum of her 20 $P_{a}$ allele values (i.e., $P_{g}$), which might be negative or positive. However, because females cannot mate with a negative number of males, $P_{p}$ is constrained to be zero or positive such that $P_{p} = 0$ when $P_{g} < 0$, but $P_{p} = P_{g}$ when $P_{g} \geq 0$. Polyandry is therefore modelled as a threshold trait \cite[][]{Lynch1998, Roff1996, Roff1998, Duthie} that is influenced by continuous genetic variation but only expressed when $P_{g} > 0$. Any negative $P_{g}$ value therefore generates phenotypic monandry, while positive $P_{g}$ values can lead to different degrees of realised polyandry (i.e., different numbers of mates, $N_{males}$).

\subsection*{Mating}

All males in the population are assumed to be available for any female to choose. We therefore assume that there is no opportunity cost of male mating, so mating with one female does not reduce a male's availability to mate with any other female. Individual females mate with their total allotment of $N_{males}$ without replacement, meaning that $N_{males}$ models a female's total number of different mates rather than solely her total number of matings.

Most often, $N_{males}$ will be smaller than the total number of available males \cite[][]{Duthie}. Each female then chooses her $N_{males}$ mates based on her pre-copulatory inbreeding avoidance phenotype ($M_{p}$), which equals her genotype ($M_{g}$). Negative or positive $M_{p}$ values cause a female to avoid or prefer mating with kin, respectively, whereas $M_{p}=0$ cause females to mate randomly with respect to kinship.

To calculate the probability that a focal female $i$ mates with a male $j$ to whom she is related by some kinship $k_{ij}$, each male is first assigned a perceived mate quality $Q^{m}_{ij}$. If the female has a strategy of pre-copulatory inbreeding avoidance ($M_{p}<0$), then $Q^{m}_{ij} = (-M_{p} \times k_{ij} + 1)^{-1}$, meaning that $Q^{m}_{ij}$ decreases linearly with increasingly positive values of $k_{ij}$ and increasingly negative values of $M_{p}$. If the female has a strategy of pre-copulatory inbreeding preference ($M_{p}>0$), then $Q^{m}_{ij} = M_{p} \times k_{ij} + 1$, meaning that $Q^{m}_{ij}$ increases with increasingly positive $k_{ij}$ and $M_{p}$. If $M_{p}=0$, then all males are assigned $Q^{m}_{ij}=1$.

Each male's value with respect to a focal female is then divided by the sum of all $Q^{m}_{ij}$ values across all males with respect to that female, thereby assigning each male a relative perceived quality $q^{m}_{ij}$, which is constrained to values between zero and one. The value of $q^{m}_{ij}$ then defines the probability that a focal female mates with the male. Mating is therefore stochastic, and females do not always mate with the male of the highest $q^{m}_{ij}$. For polyandrous females that choose multiple mates (i.e., $N_{mates}>1$), mates are chosen iteratively such that $Q^{m}_{ij}$ and $q^{m}_{ij}$ are re-calculated for each additional mate choice, and with $Q^{m}_{ij}$ and therefore $q^{m}_{ij}$ values of already chosen males set to zero to ensure mate sampling without replacement. In the unlikely event that a female's $N_{males}$ exceeds the total number of available males, then she simply mates with all males.

\subsection*{Fertilisation}

Following mating, fertilisation occurs such that each of a female's $n$ offspring is independently assigned a sire (with replacement) from the $N_{males}$ with which the female mated. Sire identity depends on female's kinship with each mate ($k_{ij}$) and her post-copulatory inbreeding strategy phenotype ($F_{p}$), which equals her genotype ($F_{g}$). Negative and positive values of $F_{p}$ correspond to post-copulatory inbreeding avoidance or preference, respectively, whereas $F_{p}=0$ causes random fertilisation with respect to kinship.

The probability that an offspring of a focal female $i$ is sired by any one of $i$'s mates $j$ is calculated by assigning a perceived fertilisation quality to each $j$ with respect to $i$, $Q^{f}_{ij}$. Perceived fertilisation quality $Q^{f}_{ij}$ is calculated in the same way as perceived mate quality $Q^{m}_{ij}$, such that if a focal female has a strategy of post-copulatory inbreeding avoidance ($F_{p}<0$), then the perceived quality of male $j$ is $Q^{f}_{ij} = (-F_{p} \times k_{ij} + 1)^{-1}$. If the focal female has strategy of post-copulatory inbreeding preference ($F_{p}>0$), then the perceived quality of male $j$ is $Q^{f}_{ij} = F_{p} \times k_{ij} + 1$. A relative quality ($q^{f}_{ij}$) is then calculated for each male by dividing his $Q^{f}_{ij}$ by the sum of the $Q^{f}_{ij}$ values across all of a female's mates. These $q^{f}_{ij}$ values, which lie between zero and one, define the probability of paternity. Females produce $n$ offspring, so $q^{f}_{ij}$ values are sampled $n$ times independently and with replacement for each female to determine the realised distribution of sires.

@@

Offsprings' alleles mutate with independent probabilities $\mu=0.001$. When a mutation occurs, a mutation effect is sampled from a normal distribution with a mean of zero and a standard deviation of $\mu_{SD}$ and added to the original allele value \cite[][]{Kimura1965, Lande1976, Bocedi2014, Duthie}. The value of $\mu_{SD}$ is set to $\sqrt{1/20}$.

@@

The viability of a focal female $i$'s offspring ($\Psi_{\textrm{off}}$) decreases as a log-linear function of her kinship with the sire $j$ of the offspring ($k_{ij}$) and inbreeding depression slope $\beta$,
\begin{equation}
\Psi_{\textrm{off}} = e^{-\beta k_{ij}}
\end{equation}
Here, $\beta$ models the number of haploid lethal equivalents that exist as deleterious recessive alleles in the gametes of $i$ and $j$, and which might be homozygous in offspring and reduce viability. Equation 1 assumes independent allelic effects, generating multiplicative effects on offspring viability \cite[][]{Morton1956, Mills1994}. We model inbreeding depression as having an absolute rather than relative effect on offspring viability (i.e., hard rather than soft selection), so that the effect of $\beta$ is consistent across generations and different parameter combinations. We assume that inbreeding always decreases offspring viability (i.e., $\beta > 0$, giving inbreeding depression but no outbreeding depression). Therefore, because $0 \leq k_{ij} \leq 1$, $-\beta \times k_{ij} \leq 0$. Values of $\Psi_{\textrm{off}}$ must therefore be between zero (if $-\beta \times k_{ij}$ is very negative) and one (if $-\beta \times k_{ij} = 0$). We therefore define $\Psi_{\textrm{off}}$ as the probability that an offspring is viable, and sample its realised viability (versus mortality) using a Bernoulli trial. Given our current objectives, we simulate evolution under conditions where inbreeding avoidance is adaptive due to strong inbreeding depression, and not evolution where inbreeding preference is adaptive due to weak or zero inbreeding depression \cite[][]{Parker1979, Kokko2006, Duthie2016a}, or given outbreeding depression \cite[][]{Bateson1983, Greeff2009}. However, as described above, positive $M_{p}$ and $F_{p}$ values resulting in inbreeding preference are permitted to evolve, and could potentially arise due to mutation or drift.

@@

After offspring mortality, $\rho$ adult immigrants are added to the focal population. The kinship between an immigrant and all other individuals always equals zero ($k_{ij}=0$). Immigration therefore prevents the mean kinship within the population from asymptoting to one over generations. To ensure that immigrants do not directly affect genotypic or phenotypic values of tendency for polyandry or pre-copulatory or post-copulatory inbreeding avoidance, immigrants are always male. Consequently, they can be chosen as females' mates based on their values of $k_{ij}=0$ but do not actively affect reproductive decisions through the expression of $P_{p}$, $M_{p}$, or $F_{p}$. Further, immigrants' $P_{a}$, $M_{a}$, and $F_{a}$ allele values are randomly sampled from normal distributions with means and standard deviations equal to those in the focal population at the time of immigration, meaning that they do not directly cause any change in the distribution of allele values. We thereby effectively assume that the focal population receives immigrants from other nearby populations that are subject to the same selection on $P_{p}$, $M_{p}$, and $F_{p}$.

@@

When post-copulatory inbreeding avoidance was allowed to evolve, mean $P_{a}$ values became substantially higher than in comparable simulations where $F_{a}$ values were fixed to zero and post-copulatory inbreeding avoidance could not evolve (Figure 1B,D,F,H). Allowing evolution of post-copulatory inbreeding avoidance alongside pre-copulatory inbreeding avoidance therefore facilitated evolution of polyandry to the degree that most females mated multiply given low costs of polyandry ($c_{P} < 0.005$; e.g., Figure 2A,B). Here, $P_{a}$ allele values increased from zero and persisted at low positive values (Figure 1B,D). Given higher costs of polyandry ($c_{P} \geq 0.005$), $P_{a}$ allele values still initially increased from zero, but then became slightly negative over generations (Figure 1F,H). Trajectories of allele values in individual simulations were typically highly stochastic, but were consistent in their long-term direction (Supporting Information Figures S1-S8). Overall, this implies that, even given strong inbreeding depression in offspring viability and simultaneous evolution of pre-copulatory and post-copulatory strategies, inbreeding avoidance might generally be a weak force driving polyandry evolution.

{\color{Gray}
    \begin{adjustwidth}{-2.2in}{0in}
    \mysidecaption{0.275}
{ 
        \begin{justify}
        \vspace{-14.25 mm}
        \captionof{figures/poly_costs.pdf}
        {\bf Figure 1:} Mean allele values underlying tendency for polyandry (red), pre-copulatory inbreeding strategy (blue), and post-copulatory inbreeding strategy (black) from simulations where post-copulatory inbreeding strategy is (A, C, E, and G) fixed to zero (i.e., random fertilisation) or (B, D, F, and H) allowed to evolve freely. Costs of polyandry ($c_p$) increase across rows from zero (A and B) to 0.01 (G and H). Mean allele values (solid lines) and associated standard errors (shading) are calculated across all individuals within a population over 40000 generations across 40 replicate populations. Negative mean allele values indicate inbreeding avoidance or tendency for monandry, and positive values indicate inbreeding preference or tendency for polyandry. Dotted lines demarcate mean allele values of zero.
    \end{justify}{\t}
}
{
   \includegraphics[height=18cm]{figures/poly_costs.pdf}%
}
    \end{adjustwidth}
}

@@

When post-copulatory inbreeding avoidance allele values ($F_{a}$) were fixed to zero, pre-copulatory inbreeding avoidance evolved even when costly (Figure 3A). Likewise, when pre-copulatory inbreeding avoidance allele values ($M_{a}$) were fixed to zero, costly post-copulatory inbreeding avoidance evolved (Figure 3C). Females therefore evolved to avoid inbreeding, and thereby avoid the cost of producing inbred offspring, through whichever route was available.

However, when both pre-copulatory and post-copulatory inbreeding avoidance could evolve, their relative evolutionary dynamics depended on their relative costs. When pre-copulatory but not post-copulatory inbreeding avoidance was costly ($c_{M}=0.02$ and $c_{F}=0$), pre-copulatory inbreeding avoidance initially evolved (i.e., negative $M_{a}$) but then evolved back towards random mating (i.e., $M_{a}=0$) following increasing evolution of post-copulatory inbreeding avoidance (Figure 3B). Similarly, when post-copulatory but not pre-copulatory inbreeding avoidance was costly ($c_{F}=0.02$ and $c_{M}=0$), post-copulatory inbreeding avoidance initially evolved ($F_{a}=0$) before evolving back to random fertilisation after ca $20000$ generations (Figure 3D).

@@

\subsection*{How do evolution of pre-copulatory and post-copulatory inbreeding avoidance affect one another?}

@@

Conversely, when pre-copulatory inbreeding allele ($M_{a}$) values were fixed for adaptive inbreeding avoidance, mean $F_{a}$ allele values did not consistently become negative over generations (Figure 4C,D). Fixation of pre-copulatory inbreeding avoidance therefore prevented evolution of post-copulatory inbreeding avoidance. However, in this case, mean $P_{a}$ allele values consistently decreased over generations, reflecting selection against polyandry regardless whether or not post-copulatory inbreeding avoidance was costly (Figure 4C,D). Consequently, when $c_{F}=0$, $F_{a}$ allele values had no effect because females were almost exclusively monandrous, resulting in high drift of $F_{a}$ values among replicates (illustrated by wide standard errors, Figure 4C). However, when $c_{F}=0.02$, $F_{a}$ values remained near zero to minimise direct costs. This lack of selection for post-copulatory inbreeding avoidance was driven by a lack of polyandry, and therefore an inability of females to bias fertilisation among multiple mates. When pre-copulatory inbreeding avoidance and polyandry were both fixed ($M_{a}=-10$ and $P_{a}=1$), $F_{a}$ allele values evolved to similarly negative means as $M_{a}$ allele values in Figure 4A,B (see Supporting Information Figure S10).

@@

This work was funded by a European Research Council Starting Grant to JMR. Computer simulations were performed using the Maxwell Computing Cluster at the University of Aberdeen.

%\begin{small}

@@

%\end{small}

>>>>

@@

We model evolution of polyandry, and of pre-copulatory and post-copulatory inbreeding avoidance strategies (hereafter simply `inbreeding strategies' because the model did not preclude evolution of inbreeding preference), in a small focal population of females and males by tracking the dynamics of alleles underlying reproductive strategies expressed by females. We thereby track evolutionary dynamics given internally consistent patterns of relatedness caused by non-random mating and allowing for effects of mutation, gene flow, drift, and selection.

All individuals have 10 physically unlinked diploid loci, and therefore 20 alleles, underlying each of three reproductive strategy traits (i.e., 30 diploid loci in total): tendency for polyandry ($P_{a}$, i.e., `polyandry' alleles), pre-copulatory inbreeding strategy ($M_{a}$, i.e., `mating' alleles), and post-copulatory inbreeding strategy ($F_{a}$, i.e., `fertilisation' alleles). All alleles can take the value of any real number \cite[i.e., a continuum-of-alleles model;][]{Kimura1965, Lande1976, Reeve2000, Bocedi2014} and combine additively to determine genotypes ($P_{g}$, $M_{g}$, and $F_{g}$) and resulting phenotypes ($P_{p}$, $M_{p}$, and $F_{p}$) for tendency for polyandry, pre-copulatory inbreeding strategy, and post-copulatory inbreeding strategy, respectively. Each individual's genotypic values $P_{g}$, $M_{g}$, and $F_{g}$ equal the sum of its 20 alleles for each trait. Each individual's phenotypic values for pre-copulatory and post-copulatory inbreeding strategy equal their respective genotypic values ($M_{p} = M_{g}$ and $F_{p} = F_{g}$), where negative and positive values cause inbreeding avoidance and preference, respectively (see details of mating and fertilisation strategies below). In contrast, phenotypic values for tendency for polyandry equal genotypic values ($P_{p} = P_{g}$) only if $P_{g} \geq 0$, while $P_{p} = 0$ if $P_{g} < 0$ because females cannot choose a negative number of mates. Polyandry is therefore modelled as a threshold trait \cite[][]{Lynch1998, Roff1996, Roff1998, Duthie} that is influenced by continuous genetic variation but only expressed when $P_{g} > 0$. A negative $P_{g}$ value therefore generates a female that is phenotypically monandrous, while a positive $P_{g}$ value generates a female that can express some degree of polyandry (see details below).

In overview, each generation proceeds with females paying costs associated with their reproductive strategy traits, and expressing polyandry, mating, and fertilisation. Offspring inherit a randomly sampled allele from each parent at each locus. Alleles can then mutate and offspring express inbreeding depression in viability. Immigrants arrive in the population and density regulation limits population growth. We record the population pedigree and directly calculate the coefficient of kinship ($k$) between all potential mates in each generation (defined as the probability that two homologous alleles will be identical-by-descent), allowing individual pre-copulatory and post-copulatory inbreeding strategies to be enacted. Key individual traits, parameter values, and variables are described in Table 1.

@@

Phenotypic values of the three reproductive strategy traits each have costs that are assumed to combine independently to increase the probability that a focal female will die before mating (see Discussion for consideration of other costs). The probabilities of pre-mating mortality due to the costs of polyandry ($c_{P}$), pre-copulatory inbreeding strategy ($c_{M}$), and post-copulatory inbreeding strategy ($c_{F}$) are $P_{p} \times c_{P}$, $|M_{p}| \times c_{M}$, and $|F_{p}| \times c_{F}$, respectively. Here $|M_{p}|$ and $|F_{p}|$ are the absolute values of $M_{p}$ and $F_{p}$, respectively. Absolute values are used for applying costs to inbreeding avoidance strategies because both negative and positive $M_{p}$ and $F_{p}$ values could potentially arise and affect the degree of inbreeding, representing inbreeding avoidance and inbreeding preference, respectively. In contrast, $P_{p}$ cannot be negative. Overall, because generations are non-overlapping, a female's probability of total reproductive failure increases linearly with the phenotypic value of each trait.

@@

\subsection*{Mating and pre-copulatory inbreeding avoidance}

After costs are realised, each remaining female chooses $N_{males}$ males to mate with, where $N_{males}$ is calculated by sampling from a Poisson distribution such that $N_{males} = Poisson(P_{p}) + 1$. This ensures that all surviving females choose at least one mate and generates each female's realised degree of polyandry with some stochastic variation around the expected mean $N_{males}$ of $P_{p}+1$.

All males in the population are assumed to be available for any female to choose. We therefore assume that there is no opportunity cost of male mating, so mating with one female does not reduce a male's availability to mate with any other female. Females mate with $N_{males}$ without replacement, meaning that $N_{males}$ models a female's total number of different mates rather than solely her total number of matings.

Most often, $N_{males}$ will be smaller than the total number of available males \cite[][]{Duthie}. Each female then chooses her $N_{males}$ mates based on her pre-copulatory inbreeding avoidance phenotype ($M_{p}$). Negative or positive $M_{p}$ values cause a female to avoid or prefer mating with kin, respectively, whereas $M_{p}=0$ cause females to mate randomly with respect to kinship.

To calculate the probability that a female $i$ mates with a male $j$ to whom she is related by some kinship $k_{ij}$, each male is first assigned a perceived mate quality $Q^{m}_{ij}$. If the female has a strategy of pre-copulatory inbreeding avoidance ($M_{p}<0$), then $Q^{m}_{ij} = (-M_{p} \times k_{ij} + 1)^{-1}$, meaning that $Q^{m}_{ij}$ decreases linearly with increasingly positive values of $k_{ij}$ and increasingly negative values of $M_{p}$. If the female has a strategy of pre-copulatory inbreeding preference ($M_{p}>0$), then $Q^{m}_{ij} = M_{p} \times k_{ij} + 1$, meaning that $Q^{m}_{ij}$ increases with increasingly positive $k_{ij}$ and $M_{p}$. If $M_{p}=0$, then all males are assigned $Q^{m}_{ij}=1$.

Each male's value with respect to a female $i$ is then divided by the sum of all $Q^{m}_{ij}$ values across all males with respect to that female, thereby assigning each male a relative perceived quality $q^{m}_{ij}$, which is constrained to values between zero and one. The value of $q^{m}_{ij}$ then defines the probability that female $i$ mates with male $j$. Mating is therefore stochastic, and females do not always mate with the male of the highest $q^{m}_{ij}$. For polyandrous females that choose multiple mates (i.e., $N_{mates}>1$), mates are chosen iteratively such that $Q^{m}_{ij}$ and $q^{m}_{ij}$ are re-calculated for each additional mate choice, and with $Q^{m}_{ij}$ and therefore $q^{m}_{ij}$ values of already chosen males set to zero to ensure mate sampling without replacement. In the unlikely event that a female's $N_{males}$ exceeds the total number of available males, then she simply mates with all males.

\subsection*{Fertilisation and post-copulatory inbreeding avoidance}

Following mating, fertilisation occurs such that each of a female's $n$ offspring is independently assigned a sire (with replacement) from the $N_{males}$ with which the female mated. Sire identity depends on female's kinship with each mate ($k_{ij}$) and her post-copulatory inbreeding strategy phenotype ($F_{p}$). Negative and positive values of $F_{p}$ correspond to post-copulatory inbreeding avoidance or preference, respectively, whereas $F_{p}=0$ causes random fertilisation with respect to kinship.

The probability that an offspring of female $i$ is sired by any one of $i$'s mates $j$ is calculated by assigning a perceived fertilisation quality to each $j$ with respect to $i$, $Q^{f}_{ij}$. Perceived fertilisation quality $Q^{f}_{ij}$ is calculated in the same way as perceived mate quality $Q^{m}_{ij}$, such that if female $i$ has a strategy of post-copulatory inbreeding avoidance ($F_{p}<0$), then the perceived quality of male $j$ is $Q^{f}_{ij} = (-F_{p} \times k_{ij} + 1)^{-1}$. If the female has strategy of post-copulatory inbreeding preference ($F_{p}>0$), then the perceived quality of male $j$ is $Q^{f}_{ij} = F_{p} \times k_{ij} + 1$. A relative quality ($q^{f}_{ij}$) is then calculated for each male by dividing his $Q^{f}_{ij}$ by the sum of the $Q^{f}_{ij}$ values across all of a female's mates. These $q^{f}_{ij}$ values, which lie between zero and one, define the probability of paternity. Females produce $n$ offspring, so a female $i$ samples from her mates $n$ times independently and with replacement with a probability of $q^{f}_{ij}$ for each male $j$ to determine the realised distribution of sires. Offspring have equal probability of being female or male.

@@

Offsprings' alleles mutate with independent probabilities $\mu=0.001$. When a mutation occurs, a mutation effect is sampled from a normal distribution with a mean of zero and a standard deviation of $\mu_{SD}$ and added to the original allele value \cite[][]{Kimura1965, Lande1976, Bocedi2014, Duthie}. The value of $\mu_{SD}$ is set to $\sqrt{1/20}$ because 20 independent alleles underlie each trait's genotypic value.

@@

The viability of a female $i$'s offspring ($\Psi_{\textrm{off}}$) decreases as a log-linear function of her kinship with the sire $j$ of the offspring ($k_{ij}$) and inbreeding depression slope $\beta$,
\begin{equation}
\Psi_{\textrm{off}} = e^{-\beta k_{ij}}
\end{equation}
Here, $\beta$ models the number of haploid lethal equivalents that exist as deleterious recessive alleles in the gametes of $i$ and $j$, and which might be homozygous in offspring and reduce viability. Equation 1 assumes independent allelic effects, generating multiplicative effects on offspring viability \cite[][]{Morton1956, Mills1994}. We model inbreeding depression as having an absolute rather than relative effect on offspring viability (i.e., hard rather than soft selection), so that the effect of $\beta$ is consistent across generations and different parameter combinations. We assume that inbreeding always decreases offspring viability (i.e., $\beta > 0$, giving inbreeding depression but no outbreeding depression). Therefore, because $0 \leq k_{ij} \leq 1$, $-\beta \times k_{ij} \leq 0$. Values of $\Psi_{\textrm{off}}$ must therefore be between zero (if $-\beta \times k_{ij}$ is very negative) and one (if $-\beta \times k_{ij} = 0$). We therefore define $\Psi_{\textrm{off}}$ as the probability that an offspring is viable, and sample its realised viability (versus mortality) using a Bernoulli trial. Given our current objectives, we simulate evolution under conditions where inbreeding avoidance is adaptive due to strong inbreeding depression, and not evolution where inbreeding preference is adaptive due to weak or zero inbreeding depression \cite[][]{Parker1979, Kokko2006, Duthie2016a}, or given outbreeding depression \cite[][]{Bateson1983, Greeff2009}. However, as described above, positive $M_{p}$ and $F_{p}$ values resulting in inbreeding preference are not precluded from evolving, and could potentially arise due to mutation or drift.

@@

After offspring mortality, $\rho$ adult immigrants are added to the focal population. The kinship between an immigrant and all other individuals always equals zero ($k_{ij}=0$). Immigration therefore prevents the mean kinship within the population from asymptoting to one over generations. To ensure that immigrants do not directly affect genotypic or phenotypic values of tendency for polyandry or pre-copulatory or post-copulatory inbreeding avoidance, immigrants are always male. Consequently, they can be chosen as females' mates based on their values of $k_{ij}=0$ but do not actively affect reproductive decisions through the expression of $P_{p}$, $M_{p}$, or $F_{p}$. Further, immigrants' $P_{a}$, $M_{a}$, and $F_{a}$ allele values are randomly sampled from normal distributions with means and standard deviations equal to those in the focal population at the time of immigration, meaning that they do not directly cause any change in the distribution of allele values. We thereby effectively assume that the focal population receives immigrants from other nearby populations that are subject to the same selection on $P_{p}$, $M_{p}$, and $F_{p}$ \cite[][]{@Duthie; @Duthie2016a}.

@@

\vspace{2 mm}
{\color{Gray}
    \begin{adjustwidth}{-2.2in}{0in}
    \mysidecaption{0.275}
{ 
        \begin{justify}
        \vspace{-14.25 mm}
        \captionof{figures/poly_costs.pdf}
        {\bf Figure 1:} Mean allele values underlying tendency for polyandry (red), pre-copulatory inbreeding strategy (blue), and post-copulatory inbreeding strategy (black) from simulations where post-copulatory inbreeding strategy is (A, C, E, and G) fixed to zero (i.e., random fertilisation) or (B, D, F, and H) allowed to evolve freely. Costs of polyandry ($c_p$) increase across rows from zero (A and B) to 0.01 (G and H). Mean allele values (solid lines) and associated standard errors (shading) are calculated across all individuals within a population over 40000 generations across 40 replicate populations. Negative mean allele values indicate inbreeding avoidance or tendency for monandry, and positive values indicate inbreeding preference or tendency for polyandry. Dotted lines demarcate mean allele values of zero.
    \end{justify}{\t}
}
{
   \includegraphics[height=18cm]{figures/poly_costs.pdf}%
}
    \end{adjustwidth}
}
\vspace{2 mm}

When post-copulatory inbreeding avoidance was allowed to evolve, mean $P_{a}$ values became substantially higher than in comparable simulations where $F_{a}$ values were fixed to zero and post-copulatory inbreeding avoidance could not evolve (Figure 1B,D,F,H). Allowing evolution of post-copulatory inbreeding avoidance alongside pre-copulatory inbreeding avoidance therefore facilitated evolution of polyandry to the degree that most females mated multiply given low costs of polyandry ($c_{P} < 0.005$; e.g., Figure 2A,B). Here, $P_{a}$ allele values increased from zero and persisted at low positive values (Figure 1B,D). Given higher costs of polyandry ($c_{P} \geq 0.005$), $P_{a}$ allele values still initially increased from zero, but then became slightly negative over generations (Figure 1F,H). Trajectories of allele values in individual simulations were typically highly stochastic, but were consistent in their long-term direction (Supporting Information Figures S1-S8). Overall, these results illustrate that post-copulatory inbreeding avoidance can facilitate evolution of polyandry as long as direct costs are sufficiently low (Figure 1B,D). However, given higher costs, evolution of polyandry is constrained even given strong inbreeding depression in offspring viability, and given resulting evolution of both pre-copulatory and post-copulatory inbreeding avoidance (Figure 1F,H).

@@

When post-copulatory inbreeding avoidance allele values ($F_{a}$) were fixed to zero, pre-copulatory inbreeding avoidance evolved even when costly (Figure 3A). Likewise, when pre-copulatory inbreeding avoidance allele values ($M_{a}$) were fixed to zero, costly post-copulatory inbreeding avoidance evolved (Figure 3C). Females therefore evolved to avoid inbreeding, and thereby avoid the indirect cost of producing inbred offspring, through whichever route was available.

However, when both pre-copulatory and post-copulatory inbreeding avoidance could evolve, their relative evolutionary dynamics depended on their relative costs. When pre-copulatory but not post-copulatory inbreeding avoidance was costly ($c_{M}=0.02$ and $c_{F}=0$), pre-copulatory inbreeding avoidance initially evolved (i.e., $M_{a} < 0$) but then evolved back towards random mating (i.e., $M_{a} \approx 0$) following increasing evolution of post-copulatory inbreeding avoidance and polyandry (Figure 3B). Similarly, when post-copulatory but not pre-copulatory inbreeding avoidance was costly ($c_{F}=0.02$ and $c_{M}=0$), post-copulatory inbreeding avoidance initially evolved (i.e., $F_{a} < 0$) before evolving back to random fertilisation (i.e., $F_{a} \approx 0$) after ca $20000$ generations (Figure 3D).

@@

\subsection*{How does fixation of pre-copulatory or post-copulatory inbreeding avoidance affect evolution?}

@@

Conversely, when pre-copulatory inbreeding allele ($M_{a}$) values were fixed for adaptive inbreeding avoidance, mean $F_{a}$ allele values did not consistently become negative over generations (Figure 4C,D). Fixation of pre-copulatory inbreeding avoidance therefore prevented evolution of post-copulatory inbreeding avoidance. In this case, mean $P_{a}$ allele values consistently decreased over generations, reflecting selection against polyandry regardless of whether or not post-copulatory inbreeding avoidance was costly (Figure 4C,D). Consequently, when $c_{F}=0$, $F_{a}$ allele values had no effect because females were almost exclusively monandrous, resulting in high drift of $F_{a}$ values (resulting in variation among replicates illustrated by the wide standard errors in Figure 4C). However, when $c_{F}=0.02$, $F_{a}$ values remained near zero to minimise direct costs. This lack of selection for post-copulatory inbreeding avoidance was driven by a lack of polyandry, and therefore an inability of females to bias fertilisation among multiple mates. When pre-copulatory inbreeding avoidance and polyandry were both fixed ($M_{a}=-10$ and $P_{a}=1$), $F_{a}$ allele values evolved to similarly negative means as $M_{a}$ allele values in Figure 4A,B (see Supporting Information Figure S10).

@@

This work was funded by a European Research Council Starting Grant to JMR. Computer simulations were performed using the Maxwell Computing Cluster at the University of Aberdeen. We thank Matthew E. Wolak for very helpful comments on this manuscript.

\begin{small}

@@

\end{small}
